---
title: "US Power Outage Map"
output: 
  html_document:
    self_contained: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

```{r load_packages, include = FALSE}
# Set repository URL
r <- getOption("repos")
r["CRAN"] <- "https://cloud.r-project.org"
options(repos = r)

# Load packages, installing as needed
if(!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(here, folders, readr, dplyr, stringr, sf, tigris, crosstalk, DT, 
               remotes)
remotes::install_github("dmurdoch/leaflet", ref = "crosstalk4", quiet = TRUE)
pacman::p_load_gh("dmurdoch/leaflet@crosstalk4")

# Set package options
options(readr.show_col_types = FALSE, readr.show_progress = FALSE)
options(tigris_use_cache = TRUE, tigris_refresh = TRUE)
```

```{r read_config_file}
# Read config file to get default folder paths, or create config file if missing
conf_file <- here('conf', 'folders.yml')
folders <- get_folders(conf_file)
```

```{r create_folders}
# Create folders, if missing
result <- create_folders(folders)
```

```{r get_outage_data}
events_csv <- here(folders$data, "outage_data.csv")
df_events <- read_csv(events_csv) %>% 
  mutate(fips = str_pad(fips, side = "left", width = 5, pad = "0"))
```

```{r prepare_map_data}
# Define the NA color used in the map (for filling polygons with NA events)
na_col_rgb <- list(r = '241', g = '240', b = '232')
na_col_hex <- with(na_col_rgb, rgb(r, g, b, maxColorValue = 255))

# Define the default color used in the map (for polygon borders)
def_col_hex <- '#BDBDC3'

# Get map county-level polygons
shape <- suppressMessages(
  counties(state = state.abb, progress_bar = FALSE, 
           cb = TRUE, resolution = '20m')) %>%
  st_as_sf() %>% st_transform('+proj=longlat +datum=WGS84')

# Combine outage data with FIPS codes by state and county name
df_map <- fips_codes %>% filter(state %in% state.abb) %>% 
  mutate(fips = paste0(state_code, county_code)) %>%
  left_join(df_events, by = c('fips'), multiple = "all") %>% 
  select(fips, 'State' = state_name, 'County' = county, 
         'Years' = n_years_available, 'Events' = total_8_hour_outages)

# Combine map data with outage data by FIPS code
shape <- shape %>% mutate(fips = paste0(STATEFP, COUNTYFP)) %>%
  left_join(df_map, by = c('fips'), multiple = "all") %>%
  select(State, County, Years, Events, geometry) %>%
  arrange(State, County)

# # Create a color palette
# mypal <- colorNumeric(palette = "YlOrRd", domain = shape$Events,
#                       na.color = na_col_hex)

# Convert Events to a factor with quantile-based bins to make a palette
event_probs <- c(0, 0.5, .75, .85, .90, .95, .98, .995, 1)
event_levels <- round(quantile(shape$Events, event_probs, na.rm = TRUE))
df_labels <- tibble(x = lag(event_levels, n = 1) + 1, y = event_levels) %>%
  mutate(x = ifelse(is.na(x), 0, x)) %>% 
  mutate(label = ifelse(x == y, as.character(x), paste(x, y, sep = "-")))
names(event_levels) <- df_labels$label
shape$EventsFct <- cut(shape$Events, breaks = c(-1, event_levels), 
                       labels = names(event_levels), ordered_result = TRUE)

# Compare Events with EventsFct using table() to check results
#table(shape$Events, useNA = 'always')
#table(shape$EventsFct, useNA = 'always')

# Create a color palette from a factor variable for filling counties
mypal <- colorFactor(palette = "magma", domain = shape$EventsFct, 
                     na.color = na_col_hex, reverse = TRUE)

# Wrap data frame in SharedData object for Crosstalk
sd <- SharedData$new(shape)
```

```{css legend_background_color}
.legend {
    background: #dddddd
}
```

The power outage map displays annual average counts of outage events lasting 8+ hours at a county level across the US Users can identify areas that experience the greatest and least burden of 8+ hour outage exposure for years between 2018-2020. Data come from PowerOutage.US, which tracks customer outage experience. This interactive map is a supplement to Do et al 2023, Spatiotemporal distribution of power outages with climate events and social vulnerability in the USA. Please refer to Do et al 2023 for further details about materials and methods and the associated GitHub for the code. 

## US Counties

```{r show_map}
# Show an interactive leaflet map
bscols(
  leaflet(sd) %>% 
    addProviderTiles("OpenStreetMap.Mapnik") %>%
  setView(lat = 37.8283, lng = -98.5795, zoom = 4) %>%
  addPolygons(
    data = sd, weight = 1, color = def_col_hex, smoothFactor = 0.2, 
    fillOpacity = 0.6, fillColor = ~ mypal(EventsFct), label = ~ County, 
    popup = ~ paste("<strong>", "# of Years of Data: ", "</strong>", Years, "<br>",
                    "<strong>", "State: ", "</strong>", State, "<br>",
                    "<strong>", "County: ", "</strong>", County, "<br>",
                    "<strong>", "Events: ", "</strong>", Events, "<br>")) %>%
  # addLegend(na.label = "NA", position = "bottomleft", pal = mypal, 
  #           values = ~ Events, title = "Annual average outages", 
  #           opacity = 0.8),
  addLegend(na.label = "NA", position = "bottomleft", pal = mypal,
            values = ~ EventsFct, title = "Annual<br>average<br>outages", 
            opacity = 0.8),
  widths = c(10)                 # Maximum width of a bscols() row is 12
)
```

```{r show_filters}
# Show check boxes for Years
bscols(
  filter_checkbox("YearsFilter", "# of Years of Data", sd, group = ~ Years,
                  inline = TRUE),
  widths = c(6)
)

# Show search boxes for the state and county
bscols(
  filter_select(id = "StateFilter", label = "State", sharedData = sd, 
                group = ~ State, multiple = TRUE),
  filter_select(id = "CountyFilter", label = "County", sharedData = sd, 
                group = ~ County, multiple = TRUE),
  widths = c(5, 5)
)
```

```{r show_table, warning=FALSE}
# Show data as interactive table
bscols(
  datatable(sd, 
            colnames = c("State" = "State", "County" = "County", 
                         "# of Years of Data" = "Years", 
                         "Annual average outages" = "Events"),
            extensions = "Scroller", style = "bootstrap", class = "stripe",
            width = "100%", height = "45%", rownames = FALSE,
            options = list(
              deferRender = TRUE, scrollY = 250, scroller = TRUE, 
              dom = 'ltipr', searching = TRUE,
              columnDefs = list(list(visible = TRUE, targets = c(0, 1, 2, 3)),
                                list(visible = FALSE, targets = '_all')))),
  widths = c(10)
)
```

## Methods and Citation
For every county, we aggregated average counts of outages lasting at least 8 hours over the county’s total number of years with data. Briefly, we defined an 8+ hour outage to be whenever the customers without power in a given county was continuously equal to or above 0.01% of customers served. We chose 0.01% based on methods used in prior studies. More information can be found in the methods section of Do et al 2023.

Citation for this map and related study:
Do, V., McBrien, H., Flores, N.M. et al. Spatiotemporal distribution of power outages with climate events and social vulnerability in the USA. Nat Commun 14, 2470 (2023). https://doi.org/10.1038/s41467-023-38084-6


For those interested in this project on power outages, we are open to collaborations and happy to discuss ways to move leverage this data to advance research. Please reach out to Dr. Joan Casey at jacasey@uw.edu for inquiries and questions.

- Source code: https://github.com/brianhigh/jacasey_power_outages

```{r}
# Cleanup unused folders
result <- cleanup_folders(folders, conf_file) 
```
